Concurrent Programming with CXL

### 2020-07-03 book/paper.tex
Latest

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be accessing a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) and free of race conditions is called \emph{thread-safe}.
\index{thread safety}%
\index{thread}%

@@

\section*{Exercises}
\begin{problems}
\item Implement an atomic swap operation.  It should take two pointer arguments
and swap the values.
\item Implement a spinlock using the atomic swap operation.
\item Write out the invariants that need to hold and check them using CXL.
\end{problems}

@@

Locks are probably the most prevalent and basic form of synchronization
in concurrent programs.  For this reason, CXL has a module called
\texttt{synch} that includes support for locks.
Figure~\ref{fig:spinlocks} shows how they are implemented, and
Figure~\ref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of Figure~\ref{fig:inc}.
Notice that the module completely hides the implementation of the
lock.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

>>>>

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be accessing a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
Typically, you want some invariant property of the data structure to hold
at the beginning and at the end of the critical section, but in the middle
the invariant may be temporarily broken---this is no issue as no other
process will be able to see it.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) and free of race conditions is called \emph{thread-safe}.
\index{thread safety}%
\index{thread}%

@@

\section*{Exercises}
\begin{problems}
\item Implement an atomic swap operation.  It should take two pointer arguments
and swap the values.
\item Implement a spinlock using the atomic swap operation.
\item Write out the invariants that need to hold and check them using CXL.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  Figure~\ref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock.
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.  Method \texttt{checker} looks for negative balances.
Running the code through CXL reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
\begin{enumerate}
\item Fix the code in Figure~\ref{fig:atm}.
\item Is it necessary to obtain a lock in \texttt{atm\_check\_balance()}?
\end{enumerate}
\end{problems}

\begin{figure}
\begin{code}
\cxlsource{atm}
\end{code}
\caption{[atm.cxl] Withdrawing money from an ATM.}
\label{fig:atm}
\end{figure}

@@

Locks are probably the most prevalent and basic form of synchronization
in concurrent programs.  Typically, whenever you have a shared data
structure, you want to protect the data structure with a lock and
acquire the lock before access and release it immediately afterward.
In other words, you want the access to the data structure to be a
critical section.
When there is a bug in a program because some code omitted obtaining
a lock before accessing the data structure, that is known as a
\emph{data race}.
\index{data race}

Because locks are so common, CXL has a module called
\texttt{synch} that includes support for locks.
Figure~\ref{fig:spinlocks} shows how they are implemented, and
Figure~\ref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of Figure~\ref{fig:inc}.
Notice that the module completely hides the implementation of the
lock.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.
